Rapid and Accurate Global PTM Discovery (G-PTM-D) Using Post-Acquisition Spectral Calibration and Defined Mass Windows

### 2017-02-28 paper.tex
changes

We have recently described a bioinformatics tool, G-PTM-D, for the discovery and localization of new PTMs from "bottom-up" tandem mass spectrometric datasets\cite{Li_2016}. The G-PTM-D workflow consists of three stages: 1) A wide-mass database search\cite{Chick_2015,Na_2011} that provides spectral matches to unmodified peptides along with the mass differences between the identified peptides and the measured parent peptide masses (hypothesized to differ in mass due to the presence of a PTM).  2) For those peptides for which the mass difference corresponds to the mass of a known PTM, a database augmentation step adds plausible localized PTMs to the proteins in the search database. 3) A final standard narrow-mass search of the augmented database to identify both modified and unmodified peptides subject to the standard FDR threshold (e.g. 1\% FDR).

Before describing the proposed changes to G-PTM-D, we introduce the database search mode definitions we use throughout. A standard \textit{narrow-mass} search considers matching theoretical peptides to spectra if the peptide mass and the identified precursor mass are equal, within some tolerance (i.e. 5ppm, or 2.1 Da). A \textit{wide-mass} search matches peptide fragment masses to fragmentation spectra while allowing large differences (i.e. $\pm 500$ Da) between the observed precursor mass and the peptide mass. An even less restrictive \textit{open-mass} search, matches peptides from a database to  fragmentation spectra without taking the precursor mass into account. We propose alternatives to these search modes, namely a \textit{notch search} and an \textit{interval search} that fall under a broad category of \textit{mass-window} searches. The notch search is an extension of the narrow-mass search: it allows exact matches (within a tolerance) to specific mass differences between the theoretical peptide and observed precursor masses. The interval search is an extension of the wide-mass search: specific problematic mass differences are excluded from allowed matches. The smaller the fraction of the mass space, and the narrower the mass window per search element, the faster the search will run (as fewer searches are executed per mass spectrum). Figure~\ref{fgr:differentSearchModes} provides a convenient way to visualize the various search modes.

@@

We recommend using the mass-window search in other procedures as well, beyond the G-PTM-D workflow. Its efficacy is demonstrated in context of a search for novel modifications, which is usually based on a wide-mass or an open-mass search. We demonstrate that a combination of specialized notch and interval searches produces superior identification and discovery results.

Along with algorithmic improvements, we recommend using spectral calibration prior to running the G-PTM-D procedure. A critical parameter for peptide identification and PTM localization is mass accuracy\cite{Scherl_2008}. Higher mass accuracy provides increased specificity and thus confidence in peptide identifications, decreasing the false discovery rate.  Instrument noise, systemic drift and miscalibration limit the mass accuracy in acquired spectra.  Multiple calibration strategies to improve mass accuracy have been devised, and fall into three general categories:  External calibration prior to the MS experiment (e.g. standard instrument calibration protocols); internal calibration during the MS experiment (e.g. real-time calibration using a lock mass standard\cite{Olsen_2005}); and subsequent to the MS experiment (post-acquisition spectral calibration, REF).  We use a post-acquisition calibration procedure, that builds upon the software lock mass concept\cite{Cox_2011} recently reported by the Mann group.  In our strategy, the m/z differences between expected and observed peaks in the peptide tandem ms spectra are compiled, and then used to recalibrate the spectra.  The increased mass accuracy of the recalibrated spectra leads directly to improved identifications of both modified and unmodified peptides, as well as to increased confidence for PTM localization.

>>>>

A recently described bioinformatics tool for the discovery and localization of new PTMs from "bottom-up" tandem mass spectrometric datasets, G-PTM-D\cite{Li_2016} is the foundation for the work described in this paper. The G-PTM-D workflow consists of three stages: 1) A wide-mass database search\cite{Chick_2015,Na_2011} that provides spectral matches to unmodified peptides along with the mass differences between the identified peptides and the measured parent peptide masses (hypothesized to differ in mass due to the presence of a PTM).  2) For those peptides for which the mass difference corresponds to the mass of a known PTM, a database augmentation step adds plausible localized PTMs to the proteins in the search database. 3) A final standard narrow-mass search of the augmented database to identify both modified and unmodified peptides subject to the standard FDR threshold (e.g. 1\% FDR).

Before describing the proposed changes to G-PTM-D, a few definitions of the various search modes must be introduced. A standard \textit{narrow-mass} search considers matching theoretical peptides to spectra if the peptide mass and the identified precursor mass are equal, within some tolerance (i.e. 5ppm, or 2.1 Da). A \textit{wide-mass} search matches peptide fragment masses to fragmentation spectra while allowing large differences (i.e. $\pm 500$ Da) between the observed precursor mass and the peptide mass. An even less restrictive \textit{open-mass} search, matches peptides from a database to  fragmentation spectra without taking the precursor mass into account. We propose alternatives to these search modes, namely a \textit{notch search} and an \textit{interval search} that fall under a broad category of \textit{mass-window} searches. The notch search is an extension of the narrow-mass search: it allows exact matches (within a tolerance) to specific mass differences between the theoretical peptide and observed precursor masses. The interval search is an extension of the wide-mass search: specific problematic mass differences are excluded from allowed matches. The smaller the fraction of the mass space, and the narrower the mass window per search element, the faster the search will run (as fewer searches are executed per mass spectrum). Figure~\ref{fgr:differentSearchModes} provides a convenient way to visualize the various search modes.

@@

The mass-window search is useful in other procedures as well, beyond the G-PTM-D workflow. Its efficacy is demonstrated in context of a search for novel modifications, which is usually based on a wide-mass or an open-mass search. A combination of specialized notch and interval searches produces superior identification and discovery results.

Along with algorithmic improvements, spectral calibration prior to running the G-PTM-D procedure is essential. A critical parameter for peptide identification and PTM localization is mass accuracy\cite{Scherl_2008}. Higher mass accuracy provides increased specificity and thus confidence in peptide identifications, decreasing the false discovery rate.  Instrument noise, systemic drift and miscalibration limit the mass accuracy in acquired spectra.  Multiple calibration strategies to improve mass accuracy have been devised, and fall into three general categories:  External calibration prior to the MS experiment (e.g. standard instrument calibration protocols); internal calibration during the MS experiment (e.g. real-time calibration using a lock mass standard\cite{Olsen_2005}); and subsequent to the MS experiment (post-acquisition spectral calibration, REF).  We use a post-acquisition calibration procedure, that builds upon the software lock mass concept\cite{Cox_2011} recently reported by the Mann group.  In our strategy, the m/z differences between expected and observed peaks in the peptide tandem ms spectra are compiled, and then used to recalibrate the spectra.  The increased mass accuracy of the recalibrated spectra leads directly to improved identifications of both modified and unmodified peptides, as well as to increased confidence for PTM localization.
